\documentclass[10pt]{article}

\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{ifthen}
\usepackage{epsfig}
\usepackage{fancyvrb}
\usepackage{url}
\usepackage{fancyhdr}

\usepackage{amsmath}
\usepackage{mathspec}
\usepackage{microtype}
\setmainfont[
    Numbers={Monospaced,OldStyle},
    BoldFont={Guardian TextEgyp Medium},
  ]{Guardian TextEgyp}\linespread{1.2}
\setsansfont[BoldFont={Helvetica Neue Medium}]{Helvetica Neue}\linespread{1.2}
\setmonofont[Scale=0.90]{Menlo}\linespread{1.2}
\setmathrm{Palatino}
\setmathfont(Digits,Latin,Greek){Palatino}

\tolerance=5000

\usepackage[margin=1.25in,includeheadfoot]{geometry}

\setlength{\skip\footins}{10mm}      % obsessing about footnote spacing
\setlength{\parskip}{1ex}            % paragraph spacing
\setlength{\parindent}{0ex}          % paragraph indentation

\usepackage[compact,noindentafter]{titlesec}
\titleformat*{\section}{\sffamily\large\bfseries}
\titlespacing{\section}{0pt}{2ex}{1ex}
\titleformat*{\subsection}{\sffamily\normalsize\bfseries}
\titlespacing{\subsection}{0pt}{2ex}{0ex}

% LISTS
\usepackage{enumitem}
\setenumerate{nolistsep,itemsep=1.0ex,parsep=0.0ex,leftmargin=*}
\setitemize{nolistsep,itemsep=1.0ex,parsep=0.0ex,leftmargin=*}

\usepackage[iso]{datetime}
\usdate

% ------- CUSTOM TITLE FORMAT -------
%
\makeatletter
\renewcommand{\maketitle}{
\begin{flushleft}          % right align
{\Large\sffamily\bfseries\@title}   % increase the font size of the title
\vspace{3ex}\\            % vertical space between the title and author name
{\normalsize\sffamily\@author}           % author name
\vspace{0ex}\\             % vertical space between author name and date
\normalsize\sffamily\@date                     % date
\vspace{5ex}              % vertical space between the author block and abstract
\end{flushleft}
}
% -----------------------------------

% DEFINE SOME COLORS
\usepackage{color}
\definecolor{light-blue}{RGB}{37,128,162}
\definecolor{light-gray}{gray}{0.975}
\definecolor{dark-gray}{gray}{0.50}

% CODE LISTINGS
\usepackage{listings}
\lstset{columns=fullflexible,
  basicstyle=\ttfamily\small,
  backgroundcolor=\color{light-gray},
  frame=lines, frame=lrtb, framesep=3pt, rulecolor=\color{dark-gray},
  aboveskip=10pt,
  belowskip=15pt,
  breaklines=true,
  showstringspaces=false,
  keepspaces=true,
}


\title{Assignment 7---Introduction to Statistics Using R}
\author{Paul Gribble}
\date{Winter, 2017}


\begin{document}
\maketitle

\thispagestyle{empty}

{\flushleft \sffamily * Due Friday Apr 14}

\section*{Bayesian Approaches: Is my coin fair?}

I give you a coin, and you flip it 10 times. The result is:

\begin{quote}
H T H H H T H T H H
\end{quote}

Answer the following questions about this result. For questions 1-4,
use an uninformative prior on $p$. Hint: remember how we saw in class
that the Beta distribution is a conjugate prior for the Binomial
likelihood distribution.

Let $p$ be the probability of obtaining H on a single flip of the
coin.

\begin{enumerate}

\item Plot the prior distribution for $p$ --- that is, for all
  possible values of $p$, plot $\mathrm{prob}(p)$.

\item Plot the likelihood distribution for $p$ --- that is, for all
  possible values of $p$, plot $\mathrm{prob}(D|p)$, where $D$ are the
  data given above for the 10 flips of the coin.

\item Plot the posterior distribution for $p$ --- that is, for all
  possible values of $p$, plot $\mathrm{prob}(p|D)$.

\item Based on the data from the 10 flips above, what is the
  most likely value of $p$?

\item What is the 95\% credible interval for $p$?

\item What is the probability that $p > 0.5$?

\end{enumerate}

\newpage
\section*{Update the model}

You decide that you want more data, so you flip the same coin another
10 times. Here is the result:

\begin{quote}
T T H T H H T H H H
\end{quote}

Incorporate these new data into your model of $p$ to answer the
following questions:

\begin{enumerate}
\setcounter{enumi}{6}

\item Plot the new prior distribution for $p$ (hint: it was the
  posterior computed above from the previous 10 flips).

\item Plot the likelihood function for $p$, given the new data.

\item Plot the new posterior distribution for $p$.

\item What is the most likely value of $p$?

\item What is the 95\% credible interval for $p$?

\item What is the probability that $p > 0.5$?

\end{enumerate}

\end{document}

