---
title: "Repeated Measures ANOVA & Mixed Designs"
output:
  html_document: default
  html_notebook: default
---

# One-way Repeated-Measures ANOVA

First let's load in the sample data. It's ``stacked'' in the usual
way, where each row is a single observation, and columns code
variables---in this case `treatment`, which has four levels,
and `subject`, which has 10 levels.

```{r}
datafilename <- "http://www.gribblelab.org/stats/data/oneWayRepdata.csv"
mydata <- read.table(datafilename, sep=",", header=TRUE)
mydata$treatment <- factor(mydata$treatment)
mydata$subject <- factor(mydata$subject)
mydata
```

## demo: ignore subjects

As a demonstration, let's run this *as if* it were simply a
between-subjects ANOVA, in other words, ignoring the fact that
observations from the four levels of `treatment` come from the
same subjects:

```{r}
am1 <- aov(dv ~ treatment, data=mydata)
summary(am1)
```

In this case the F-test of the `treatment` effect is
significant at $p=0.001890$.

## demo: include subjects as if it were a two-factor between-subjects design

Now as another demonstration, let's run this ANOVA again *as if*
it were a two-factor design, in which we include `subject` as a
factor. In this case, note that we cannot include both the main
effects *and* the interaction effect---since there is only one
observation from each subject in each condition, there are not enough
degrees of freedom. So we will simply specify a model that has the two
main effects and leaves out the interaction effect:

```{r}
am2 <- aov(dv ~ treatment + subject, data=mydata)
summary(am2)
```

Note how the SS, df, and MS for the `treatment` effect are
identical to the previous case where the `subject` factor was
not included in the model. What's changed, is that the *error term* (the term denoted `Residuals`) is now smaller. In other
words, in the first model (`am1`), the SS was $77.00$ --- this
represents the ``leftover'' variability in the dependent variable that
is unaccounted for by the `treatment` factor. In the
`am2` model, the `subject` factor has a SS of $48.4$,
which is now sliced out of the error term, leaving only $28.6$. In
other words, the `subject` factor has accounted for a chunk of
the variance in the dependent variable, leaving us with a smaller
error term in the ANOVA. This is good, because it means that the
F-ratio for the `treatment` effect is larger (and hence the
p-value is smaller)---since it is equal to the MS for
`treatment` divided by the MS for the error term.

Note that this decrease in MS error is not ``free'', in other words
there is a cost: we give up degrees of freedom (was $36$ in
`am1` but is now reduced to $27$ in `am2`). Lower
degrees of freedom in the error term means that (all other things
being equal) the MS error term will be larger, which (all other things
being equal) means a smaller F-ratio for our test of the
`treatment` effect. Now all other things are *not* equal,
because the inclusion of `subjects` has lowered the error term
SS as well ... so whether the inclusion of additional factors helps
us, in the end, or not, ultimately depends on the usual question: is
the reduction in error gained by the inclusion of the additional term
in the model *worth it*, given that we have to "pay" a certain
number of degrees of freedom?

## Correct method (univariate approach)

Now the correct way of running a repeated-measures ANOVA is not to
pretend that this is a 2-factor between-subjects design, minus the
interaction term---(even though in the end, the computations of the
ANOVA table are the same). Here is the correct way of doing it:

```{r}
am3 <- aov(dv ~ treatment + Error(subject/treatment), data=mydata)
summary(am3)
```

Note that the specification of the model has a new element we have not
seen before, namely the addition of an error term:
`+Error(subject/treatment)`. This notation tells R to slice out
of the model an additional error term corresponding to the variance
accounted for by subjects. The `subjects/treatment` notation
tells R that the `treatment` factor is the repeated-measures
factor over which subjects applies.

As you can see the output is essentially the same, just reorganized a
bit. The variance accounted for by `subjects` has not only been
removed from the model's overall error term, but has even been
physically separated in the output. So now, the effect of
`treatment` is significant at $p=3.06e-05$.

This is the correct method of running a single factor
repeated-measures ANOVA, using what is called a *univariate approach*. That is to say, there is a single dependent variable
(called `dv` in our data frame `mydata`). We have been
doing univariate procedures all along in the course so far.

## Multivariate approach

There is another general method of performing ANOVA when there are one
or more repeated-measures factors, called the multivariate
approach. Maxwell & Delaney devote a lot of material to going over
the rationale. In short, although it is slightly more complex
conceptually, one benefit of using a multivariate approach is that it
provides ways of testing the *sphericity assumptio*---the
assumption of homogeneity of variance of differences between groups---as well as providing corrected versions of the F-test assuming the
assumption has been violated. As you know from your reading,
repeated-measures ANOVA is actually quite sensitive to the sphericity
assumption, and what's more, the sphericity assumption is very often
violated.

Here is how to perform a repeated-measures ANOVA in R using a
multivariate approach. First, we must reorganize the data into a
format in which each row represents a single subject, and columns
represent levels of the `treatment` factor. This may be
familiar to those of you who have used SPSS---this is the way
repeated-measures factors are organized in SPSS data tables as well.

```{r}
response <- with(mydata,cbind(dv[treatment==1], dv[treatment==2],
	dv[treatment==3], dv[treatment==4]))
```

Our data are now essentially in a matrix format:

```{r}
response
```

We now form the multivariate model using the `lm()` function in R:

```{r}
mlm1 <- lm(response ~ 1)
mlm1
```

The `~1` notation simply tells R that there are no
between-subjects factors here ... in other words, only fit the model
using intercepts. If you type `mlm1` to look at the model
object you will see that four intercepts were fit---one representing
the mean of each of the four levels of the dependent variable:

Now we must set up a variable that defines the *design* of our
study, which is simple in this case, it's a single factor with four
levels:

```{r}
rfactor <- factor(c("r1", "r2", "r3", "r4"))
```

Now we must load the `car` library into R (we only have to do
this once in a given session), because we are going to make use of the
`Anova()` function (note the capital A, this is
different than the `anova()` function). The `Anova()`
function calculates ANOVA tables for a number of different kinds of
model objects including multivariate model objects.

```{r}
library(car)
```

We now define a new anova model object `mlm1.aov` by a function
call to `Anova()`:

```{r}
mlm1.aov <- Anova(mlm1, idata=data.frame(rfactor), idesign = ~rfactor, type="III")
```

The first argument, `mlm1`, is our multivariate model defined
above. The second argument, `idata=data.frame(rfactor)` passes
information about the within-subjects variable, in other words how
many levels there are. The third argument, `idesign=~rfactor`,
passes information about the within-subjects design, in other words
that the variable that `rfactor` describes is the
repeated-measures variable. The fourth argument, `type="III"`,
instructs `Anova()` to calculate the "Type-III" sums of
squares when forming the ANOVA table. This only matters for so-called
"unbalanced designs" in which there are different numbers of
observations in different groups. Maxwell and Delaney do a good job of
explaining how this is related to different ways of computing sums of
squares. In the `summary()` command we specify
`multivariate=FALSE` because we want to suppress the portion of
the output of the `Anova()` function that is related to
multivariate statistical tests---these are really only relevant when
the experimental design truly is a multivariate design (in other words
when there are multiple dependent variables).

Here is the output we get:

```{r}
summary(mlm1.aov, multivariate=FALSE)
```

First we see that the test of the repeated measures factor (called
`rfactor` here) is significant at $p=3.06e-05$. Then we see
that a test of sphericity was performed (Mauchly's test), which is not
significant ($p=0.14884$). Nevertheless, we get two corrected versions
of the F-test: one called Greenhouse-Geisser, and another, slightly
less conservative correction, called Huynh-Feldt. In this case, both
are also significant (the p-values for GG and HF are marked
`Pr(>F[GG])` and `Pr(>F[HF])`, respectively). See your
Maxwell and Delaney readings for details about how these corrections
are computed.

## Comparisons between individual means

There are two legitimate ways of testing differences between
individual means. The first is to use the same F-ratio that we have
seen from the between-subjects ANOVA, for testing contrasts $\psi$:

\begin{equation}
	F_{comp} = \frac{\psi^{2}/\sum_{j=1}^{a}(c^{2}_{j}/n_{j})}{MS_{error}}
\end{equation}

where $\psi$ is the contrast of interest:

\begin{equation}
	\psi = \sum_{j=1}^{a}(c_{j}\bar{Y}_{j})
\end{equation}

So for example if we want to test the mean of group 1 minus the mean
of group 2, the weights on the contrast would be $c_{1}=(+1)$ and
$c_{2}=(-1)$. We would simply compute the $MS_{error}$ term based on
the ANOVA output. Remember, $MS_{error}=SS_{error}/df_{error}$:

```{r}
df2 <- 27 # df for error term
sserr <- 28.6 # ss error term
mserr <- sserr/df2 # compute mserr
n <- 10 # num subjects per group
a <- 4 # num groups
mresp1 <- mean(response[,1]) # mean of resp1
mresp2 <- mean(response[,2]) # mean of resp2
sscomp <- (mresp1-mresp2)^2 # ss comparison
dfcomp <- 1 # df comparison
Fcomp <- (n*sscomp/2)/(mserr) # Fobs for comparison
Fcomp
```

We can then compute a probability using the `pf()` function:

```{r}
pcomp <- 1-pf(Fcomp, dfcomp, df2) # pobs for comparison
pcomp
```

In this case, the mean of group 1 is not significantly different than
the mean of group 2, $p=0.1399$.  Note that this is a pairwise
comparison that is uncorrected for Type-I error. We can, for example,
perform this as a Tukey test, by transforming $Fcomp$ into a value of
$q$ (studentized range statistic), and then computing a probability
(see Maxwell & Delaney, pg. 550, equation 41)

```{r}
qobs <- sqrt(2*Fcomp) # compute q value
pt <- 1 - ptukey(qobs, 4, (4-1)*(n-1)) # df for q are (a,(a-1)(n-1))
pt
```

Again, mean 1 is not significantly different than mean 2, $p=0.5605$.

### If you're concerned about sphericity

If you're concerned about violating the sphericity assumption
(heterogeneity of variances of differences between groups), then you
might notice that the method above is based on using the $MS_{error}$
term from the ANOVA for the denominator of the F-test. Essentially
this is a sort of pooled estimate of variance based on all groups. If
sphericity is violated, then it may be a better approach to perform a
pairwise test in a way that only uses variances of the groups of
interest to perform the test. One way to do this is by forming a
contrast, based on a set of weights (as above), but that is applied
not to the means of each group, but to individual subject scores, to
create composite scores $\psi_{i}$:

\begin{equation}
	\psi_{i} = \sum_{j=1}^{a}c_{j}Y_{ij}
\end{equation}

Then the values of the composite scores $\psi_{i}$ are evaluated using
a t-test. The extent to which they are non-zero reflects the extent to
which the contrast is significant. In other words the t-test tests the
hypothesis that the composite scores were drawn from a zero-mean
population. Here is an example, again comparing group 1 vs group 2:

```{r}
# fancy matrix multiplication way of computing composite scores:
mycontrast <- c(+1, -1, 0, 0)
dscores <- response %*% mycontrast
# or you can simply do this:
dscores <- response[,1]-response[,2]
t.test(dscores)
```

So the difference between groups 1 and 2 is not significant
($p=0.1323$). Note that this test is uncorrected for Type-I error. You
could always transform a value of the $t$ statistic into an F (when
$df_{num}=1$, $F=t^{2}$), then into a $q$ statistic (as above) to
perform it as a Tukey test.


