---
title: "Split-Plot ANOVA"
output:
  html_document: default
  html_notebook: default
---

# Higher order repeated measures designs & MANOVA

## Two factor repeated measures ANOVA

First let's load in the sample data. We have organized the data file in a column-wise format, where each row is a
subject. This is in preparation for a multivariate approach to the ANOVA. Here we won't both with the univariate
approach, since we are interested in the sphericity tests and the corrected values of the F test.

```{r}
fname <- "http://www.gribblelab.org/stats/data/2wayrepdata.csv"
mydata <- read.table(fname, sep=",", header=TRUE)
mydata
```

Let's extract the data matrix (just the numeric values) from the data frame:

```{r}
dm <- as.matrix(mydata[1:5, 2:7])
dm
```

Now let's create a multivariate linear model object:

```{r}
mlm1 <- lm(dm ~ 1)
```

Next we are going to use the `Anova()` command in the `car` package, so we have to first load the
package. We also have to define a data frame that contains the within-subjects factors. If you don't have the `car` package installed, just type `install.packages("car")` and R will download it and install it.

```{r}
library(car)
af <- factor(c("a1","a1","a1","a2","a2","a2"))
bf <- factor(c("b1","b2","b3","b1","b2","b3"))
myfac <- data.frame(factorA=af,factorB=bf)
myfac
```

Now we will define the anova using `Anova()`:

```{r}
mlm1.aov <- Anova(mlm1, idata = myfac, idesign = ~factorA*factorB, type="III")
summary(mlm1.aov, multivariate=FALSE)
```

Note how the formula \texttt{FactorA*FactorB} uses the same names as in the `myfac` data frame we defined.

The first part of the output lists the omnibus F tests for the main effects and the interaction effect. We then see
the Mauchly tests of sphericity. We see tests for the main effect of factorB, and the factorA:factorB interaction
effect. We don't see a test of the main effect of factorA, because in this case, factorA has only two levels ...
and so there is no variances of differences-between-groups... since there are only two levels, there is only a
single variance of differences (between the two levels). We then see Greenhouse-Geisser and Huynh-Felft
corrections.

### Simple main effects

The factorA:factorB interaction is significant, so we want to conduct so-called simple main effects analyses. This
would be testing the effects of one factor (e.g. factorB), separately within each level of factorA (or vice-versa).
In a between-subjects two-factor ANOVA, simple main effects are evaluated by doing separate one-way ANOVAs, but
using the MSerror term from the overall two-factor analysis as the error term. For within-subjects designs it's
probably better to use separate error terms for each analysis, since the sphericity assumption is likely not true, and repeated measures ANOVA Is sensitive (more so than between-subjects ANOVA) to violations of the sphericity assumption.
Therefore we can in fact literally run separate single-factor repeated measures ANOVAs, with one factor, within
levels of the other factor.

### Pairwise tests & linear contrasts

The approach for computing linear contrasts (including pairwise tests) is the same as for a single-factor repeated
measures design. We can either compute F ratios by taking the appropriate MSerr term from the ANOVA output (this
approach assumes sphericity), or we can simple compute differences scores and perform t-tests (this doesn't assume
sphericity). Correcting for Type-I error is up to you --- you could use a Bonferroni adjustment, or compute Tukey
probabilities, etc.

## Split plot designs

A split plot design is a mixed design in which there are some repeated measures factor(s) and some between-subjects
factor(s). Let's load in some sample data for a study with one repeated measures and one between subjects factor:

```{r}
fname <- "http://www.gribblelab.org/stats/data/splitplotdata.csv"
mdata <- read.table(fname, sep=",", header=TRUE)
mdata
```

We have three levels of a repeated measures factor (a1, a2, a3) and two levels of a between-subjects factor, gender
(m,f), and 10 subjects.

First as before we extract the data corresponding to the dependent variable from the data frame:

```{r}
dm <- as.matrix(mdata[1:10, 2:4])
dm
```

Then we formulate our multivariate model:

```{r}
mlm <- lm(dm ~ 1 + gender, data=mdata)
```

Note how now `dm` depends not just on a constant but also on `gender`. Next we design a data frame
that contains the design of the repeated measures factor:

```{r}
af <- factor(c("a1","a2","a3")) 
myfac <- data.frame(factorA=af) 
myfac
```

Now we use the `Anova()` function to perform the split plot anova:

```{r}
mlm.aov <- Anova(mlm, idata=myfac, idesign = ~factorA, type="III")
summary(mlm.aov, multivariate=FALSE)
```

### Followup tests

The rules and approach for further tests following significant omnibus ANOVA test(s) are no different than before.


